---
title: "Singapore Birth and Fertility EDA"
author: "Ebony Hunt"
date: "2026-03-19"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
header-includes:
  \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyr)
library(dplyr)
library(tsibble)
library(feasts)
library(fable)
library(ggplot2)
library(forecast)
```

### GitHub Repository:

### <https://github.com/a1926113/Time_Series_Analysis>

Firstly, the data is read in. This is a tsibble with data collected from 1960 - 2025 detailing the annual birth and fertility rates in Singapore, collected by the Singapore Department of Statistics. 

```{r read_data, warning=FALSE, show_col_types = FALSE}
BirthsAndFertilityRatesAnnual <- 
  read_csv("C:/Users/ebony/OneDrive - University of Adelaide/Desktop/2026 Time Series/BirthsAndFertilityRatesAnnual (1).csv")
```

There are some years that are character variables, and some years that are doubles. Every year is converted to doubles and data listed as "na" is transformed into correctly formatted NAs.

```{r to_double}

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual |>
  mutate(across(2:67, as.character)) |>
  mutate(across(2:67, ~na_if(., "na"))) |>
  mutate(across(2:67, as.double))

```

The data is not in an optimal format, with each year listed as a column. The data is transformed so that the years become the rows and the variables become the columns. The report is only looking to investigate the Total Fertility Rate, and the Total Live Births so every other column will also be filtered out.

```{r pivot_longer,warning=FALSE}

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual |>
  pivot_longer(cols = -DataSeries,
               names_to = "Year",
               values_to = "Value") |>
  pivot_wider(names_from = DataSeries,
              values_from = "Value")

BirthsAndFertilityRatesAnnual$Year <-as.integer(BirthsAndFertilityRatesAnnual$"Year")

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual[, -c(3:15, 17,18)]

head(BirthsAndFertilityRatesAnnual)
```

The data must be changed into a tsibble for further analysis. The variables are also given the more appropriate names TFR, and TLB.

```{r}

BirthFertilityRates<- as_tsibble(BirthsAndFertilityRatesAnnual,
                  index = Year) |>
  rename(year = Year, TFR = "Total Fertility Rate (TFR)",
         TLB = "Total Live-Births") 

head(BirthFertilityRates)

```
For the purpose of this report, the data from 1960-2012 will be used to create time series models for prediction of the total live births and total fertility rate from 2013-2024. 

The data will be split into a new tsibble called BFR1 with the data from these date ranges. 

```{r}
BFR1 <- BirthFertilityRates[1:53,]
```

Plots of the total fertility rate, and total live births per year in Singapore are visualized to get an overview of the data and identify patterns that will need to be considered when fitting the final models  for prediction.

Figure 1 shows a non seasonal, non linear decreasing trend of the total fertility rate in Singapore from 1960 to 2012.


```{r, warning=FALSE, fig.cap = "Time plot of total fertility rate for Singapore per year from 1960-2012", out.width="50%", fig.align="center",fig.show='asis'}
BFR1 |>
  autoplot(TFR) +
  labs(x = "Year",
       y = "Total Fertility Rate", 
       title = "Total Fertility Rate Per Year in Singapore")
```
\FloatBarrier

The plot of the total live births per year in Singapore in Figure 2 shows a non seasonal decreasing trend. There is a large spike from 1986 to 1987 that should be investigated, there may be a long term cyclic component that is not currently obvious from range of dates observed. 

```{r, warning=FALSE, fig.cap="Time plot of total live births for Singapore per year from 1960-2012", out.width="50%", fig.align="center", fig.show='asis'}
BFR1 |>
  autoplot(TLB) +
  labs(x = "Year",
       y = "Total Live Births",
       title = "Total Live Births Per Year in Singapore")
```
\FloatBarrier

The total fertility birth per year in Singapore will be considered first. Before fitting an appropriate time series model the data will need to be made linear for simplicity. This will be achieved by doing a log-transform on the data for the total fertility rate.

Figure 3 now shows an approximately linear downwards trend. There is also a more obvious jump in the total fertility rate during the mid 1980's, around a similar time that figure 2 showed a sharp increase in total live births. It may be beneficial to investigate additional factors such as immigration rates and economic influences to determine a possible cause for this spike. 

```{r, warning=FALSE, fig.cap = "Time plot of the log-transformed fertility rate in Singapore from 1960-2012", out.width="50%", fig.align="center",fig.show='asis'}

BFR1 <- BFR1 |>
  mutate(TFR.log = log(TFR))

BFR1 |>
  autoplot(TFR.log) +
  labs(x = "Year",
       y = "Total Fertility Rate", 
       title = "Total Fertility Rate Per Year in Singapore")
```
