---
title: "Singapore Birth and Fertility EDA"
author: "Ebony Hunt"
date: "2026-03-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyr)
library(dplyr)
library(tsibble)
library(feasts)
library(fable)
library(ggplot2)
```

### github Repository: <https://github.com/a1926113/Time_Series_Analysis>

Read in the data

```{r read_data, warning=FALSE, show_col_types = FALSE}
BirthsAndFertilityRatesAnnual <- read_csv("C:/Users/ebony/OneDrive - University of Adelaide/Desktop/2026 Time Series/BirthsAndFertilityRatesAnnual (1).csv")

head(BirthsAndFertilityRatesAnnual)
```
There are some years that are character variables, and some years that are doubles. Every year is converted to doubles and data listed as "na" is reformatted to be a correct NA.
```{r to_double}

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual |>
  mutate(across(2:67, as.character)) |>
  mutate(across(2:67, ~na_if(., "na"))) |>
  mutate(across(2:67, as.double))

BirthsAndFertilityRatesAnnual$Year <- as.integer(BirthsAndFertilityRatesAnnual$Year)

``` 
The data is not in an optimal format. The years become the rows and the variables become the columns. The report is only looking to investigate the Total Fertility Rate, and the Total Live Births so every other column will also be filtered out. 

```{r pivot_longer,warning=FALSE}

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual |>
  pivot_longer(cols = -DataSeries,
               names_to = "Year",
               values_to = "Value") |>
  pivot_wider(names_from = DataSeries,
              values_from = "Value")

head(BirthsAndFertilityRatesAnnual)

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual[, -c(3:15, 17,18)]

head(BirthsAndFertilityRatesAnnual)
```
The data must be changed into a tsibble for further analysis.
The variables are also given more appropriate names.

```{r}

BirthFertilityRates<- as_tsibble(BirthsAndFertilityRatesAnnual,
                  index = Year) |>
  rename(year = Year, TFR = "Total Fertility Rate (TFR)",
         TLB = "Total Live-Births") 

head(BirthFertilityRates)

```