---
title: "Singapore Birth and Fertility EDA"
author: "Ebony Hunt"
date: "2026-03-19"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
header-includes:
  \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyr)
library(dplyr)
library(tsibble)
library(feasts)
library(fable)
library(ggplot2)
library(forecast)
```

### GitHub Repository:

### <https://github.com/a1926113/Time_Series_Analysis>

Firstly, the data is read in. This is a tsibble with data collected from 1960 - 2025 detailing the annual birth and fertility rates in Singapore, collected by the Singapore Department of Statistics.

```{r read_data, warning=FALSE, show_col_types = FALSE}
BirthsAndFertilityRatesAnnual <- 
  read_csv("C:/Users/ebony/OneDrive - University of Adelaide/Desktop/2026 Time Series/BirthsAndFertilityRatesAnnual (1).csv")
```

There are some years that are character variables, and some years that are doubles. Every year is converted to doubles and data listed as "na" is transformed into correctly formatted NAs.

```{r to_double}

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual |>
  mutate(across(2:67, as.character)) |>
  mutate(across(2:67, ~na_if(., "na"))) |>
  mutate(across(2:67, as.double))

```

The data is not in an optimal format, with each year listed as a column. The data is transformed so that the years become the rows and the variables become the columns. The report is only looking to investigate the Total Fertility Rate, and the Total Live Births so every other column will also be filtered out.

```{r pivot_longer,warning=FALSE}

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual |>
  pivot_longer(cols = -DataSeries,
               names_to = "Year",
               values_to = "Value") |>
  pivot_wider(names_from = DataSeries,
              values_from = "Value")

BirthsAndFertilityRatesAnnual$Year <-as.integer(BirthsAndFertilityRatesAnnual$"Year")

BirthsAndFertilityRatesAnnual <- BirthsAndFertilityRatesAnnual[, -c(3:15, 17,18)]

head(BirthsAndFertilityRatesAnnual)
```

For the purpose of this report, the data from 1960-2012 will be used to create time series models for prediction of the total live births and total fertility rate from 2013-2024. However, for the exploratory data analysis the data set from 1960-2024 will be considered. The new tibble with these dates will be created and called BFR1 for simplicity

```{r}

BFR <- BirthsAndFertilityRatesAnnual[2:66, ]

```

The data must be changed into a tsibble for further analysis. The variables are also given the more appropriate names TFR, and TLB. Additionally, the tsibble is renamed BFR1 for simplicity

```{r}

BFR  <-
  as_tsibble(BFR,
             index = Year) |>
  rename(year = Year, TFR = "Total Fertility Rate (TFR)",
         TLB = "Total Live-Births") 

head(BFR)

```

Plots of the total fertility rate, and total live births per year in Singapore are visualized to get an overview of the data and identify patterns that will need to be considered when fitting the final models for prediction.

Figure 1 shows a non seasonal, non linear decreasing trend of the total fertility rate in Singapore from 1960 to 2024.

```{r, warning=FALSE, fig.cap = "Time plot of total fertility rate for Singapore per year from 1960-2024", out.width="50%", fig.align="center",fig.show='asis'}
BFR |>
  autoplot(TFR) +
  labs(x = "Year",
       y = "Total Fertility Rate", 
       title = "Total Fertility Rate Per Year in Singapore")
```

\FloatBarrier

Doing a portmanteau test for the Total Live Birth rate shows a p value of 1.623146e-13 indicating that the model likely have at least one autocorrelated residual. The result of the portmanteau test and clear evidence of a trend in Figure 1 indicate that the data is not stationary.

```{r}

BFR |> features(TFR, portmanteau_tests)

```

The plot of the total live births per year in Singapore in Figure 2 shows a non seasonal decreasing trend. There is a large spike from 1986 to 1987 that should be investigated, there may be a long term cyclic component that is not currently obvious from range of dates observed.

```{r, warning=FALSE, fig.cap="Time plot of total live births for Singapore per year from 1960-2024", out.width="50%", fig.align="center", fig.show='asis'}
BFR |>
  autoplot(TLB) +
  labs(x = "Year",
       y = "Total Live Births",
       title = "Total Live Births Per Year in Singapore")
```

\FloatBarrier

A portmanteau test of using the Total Live Births data produced a p-value of 7.66609e-12 indicating that it is highly likely there are autocorrelations between the residuals at at least one lag. Similarly to thhe Total Fertility rate data,the decreasing trend present in the time plot and the portmanteau test indicate that the data is nonstationary.

```{r}

BFR |> features(TLB, portmanteau_tests)

```

For the purposes of the exploratory data analysis, models will be fit to the years 1960-2012. The new tsibble with these dates will be given the name BFR1

```{r}

BFR1 <- BFR[1:53, ]

```

The total fertility rate per year in Singapore will be considered first. Before fitting an appropriate time series model the data will need to be made linear for simplicity. This will be achieved by doing a log-transform on the data for the total fertility rate.

Figure 3 now appears to show an approximately linear downwards trend. There is also a more obvious jump in the total fertility rate during the mid 1980's, around a similar time that Figure 2 showed a sharp increase in total live births. It may be beneficial to investigate additional factors such as immigration rates and economic influences to determine a possible cause for this spike.

```{r, warning=FALSE, fig.cap = "Time plot of the log-transformed fertility rate in Singapore from 1960-2012", out.width="50%", fig.align="center",fig.show='asis'}

BFR1 <- BFR1 |>
  mutate(TFR.log = log(TFR))

BFR1 |>
  autoplot(TFR.log) +
  labs(x = "Year",
       y = "Total Log Transformed Fertility Rate", 
       title = "Total Fertility Rate Per Year in Singapore")
```

The log transformed data will now be differenced to attempt to detrend the data. Figure 4 shows the data with a less obvious trend, appearing more stationary. The spike present in the data during the late 1980's is more pronounced. 

```{r, fig.cap = "Time plot of the log-transformed differenced fertility rate in Singapore from 1960-2012", out.width="50%", fig.align="center",fig.show='asis'}

BFR1 <- BFR1 |> mutate(D.TFR.log = 
                         difference(TFR.log, lag = 1))

BFR1 |> autoplot(D.TFR.log) +
  labs(x = "Year",
       y = "Total long Transformed Differenced Fertility Rate", 
       title = "Total Fertility Rate Per Year in Singapore")

```

\FloatBarrier

An auto regressive model will be fit to the transformed and differenced data.

```{r}
TFR.ar <- BFR1 |> 
  # remove NAs from differencing before model
  filter(!is.na(D.TFR.log)) |>
  model(
    D.ar = AR(D.TFR.log ~ order(1))
  )

tidy(TFR.ar)

```

The ACF plot in Figure 5 shows a significant lags the 5% level of significance at lag 12 and 13. As the confidence level is 5% it may be reasonable to expect that the plot shows one significant lag, however the presence of more than one of spikes and the magnitude of the spike at lag 12 suggests autocorrelation between lags.

```{r, fig.cap = "ACF plot of the Total Fertility Rate autoregressive model", out.width="50%", fig.align="center",fig.show='asis'}

augment(TFR.ar) |>
  ACF(.resid) |> 
  autoplot()

```

\FloatBarrier

The data will be differenced again with lag 24 to attempt to remove the autocorrelations between the residuals 
An arima model will be fit to help address these autocorrelations.

Figure 6 shows no significant lags at the 5% significance level. This indicates that the residuals are white noise. 

```{r fig.cap = "ACF plot of the Total Fertility Rate autoregressive moving average model", out.width="50%", fig.align="center",fig.show='asis'}

BFR1 <- BFR1 |> mutate(SD.TFR.log = 
                         difference(D.TFR.log, lag = 24))

TFR.arima <- BFR1 |> 
  model( arima1 = ARIMA(SD.TFR.log ~ pdq(1,0,0)
                        + PDQ(0,1,0)))

augment(TFR.arima) |>
  ACF(.resid) |> 
  autoplot()

```

\FloatBarrier

checking the Akaike Information Criterion (AIC) shows that the ARIMA model has a lower value confirming it a more accurate model for prediction. 

```{r}

glance(TFR.arima)

glance(TFR.ar)

```

The Total Live Births per year in Singapore from 1960-2012 will now be considered. Before fitting an appropriate time series model the data will need to be differenced to attempt to detrend the time plot.

Figure 7 shows the data with a less obvious trend, appearing more stationary. Similarly to the Fertility Rate time plot, the spike present in the data during the late 1980's is more pronounced. 


```{r, fig.cap = "Time plot of the differenced Live Births in Singapore from 1960-2012", out.width="50%", fig.align="center",fig.show='asis'}

BFR1 <- BFR1 |> mutate(D.TLB = 
                         difference(TLB, lag = 1))

BFR1 |> autoplot(D.TLB) +
  labs(x = "Year",
       y = "Total Differenced Live Births", 
       title = "Total Differenced Live Births Per Year in Singapore")

```

\FloatBarrier

An auto regressive model will be fit to the differenced data.

```{r}

TLB.ar <- BFR1 |>
  filter(!is.na(D.TLB)) |>
  model(D.ar1 = AR(D.TLB ~ order(1)))

tidy(TLB.ar)

```

The ACF plot in Figure 8 shows a significant lags the 5% level of significance at lags 12 and 13 Suggesting autocorrelation between lags.

```{r, fig.cap = "ACF plot of the Total Live Birth autoregressive model", out.width="50%", fig.align="center",fig.show='asis'}

augment(TLB.ar) |>
  ACF(.resid) |> 
  autoplot()

```

\FloatBarrier

Using the same technique that was used to refine the total fertility rate model he data will be differenced again with lag 24 to attempt to remove the autocorrelations between the residuals.
An arima model will be fit to help address these autocorrelations.

Figure 9 shows no significant lags at the 5% significance level. This indicates that the residuals are white noise. 

```{r, fig.cap = "ACF plot of the Total Live Birth autoregressive moving average model", out.width="50%", fig.align="center",fig.show='asis'}

BFR1 <- BFR1 |> mutate(SD.TLB = 
                         difference(D.TLB, lag = 24))

TLB.arima <- BFR1 |> 
  model( arima1 = ARIMA(SD.TLB ~ pdq(1,0,0)
                        + PDQ(0,1,0)))

augment(TLB.arima) |>
  ACF(.resid) |> 
  autoplot()

```

\FloatBarrier

Checking the Akaike Information Criterion (AIC) shows that the ARIMA model has a significantly higher value meaning it is not as effective for prediction as the autoregressive model despite the fact there are no autocorrelated lags. 

```{r}

glance(TLB.arima)

glance(TLB.ar)

```

In the final report, when more than one model has been created for each time series, there are a variety of methods that can be used to test which model is best for prediction. The Akaike Information Criterion (AIC) or the Ljung-Box test for each model can be compared, with the lower value being the preferred model. Additionally, the root mean squared error of the predictions can be tested to see how much they differ from the data recorded from 2013-2024. A visual inspection can also useful to assess the accuracy of the predictions. 